### 2012-09-18 doc/prelim/document/chapters/linear_problem/linear_problem.tex
aworking on parallel matrix multiplication

We note here several properties of this formulation and how they may
facilitate or hinder the solution of large-scale, sparse linear
problems, also noting that these properties are common among many
Krylov methods. First, from a memory perspective GMRES is efficient in
that the operator $\ve{A}$ need not be explicitly stored. Rather, only
the ability to compute the action of that operator on a vector of
valid size is required. However, these savings in memory are balanced
by the fact that the long recurrence relations used in the Arnoldi
procedure require all vectors that span the Krylov space to be
stored. If the size of these vectors becomes prohibitive, the Arnoldi
procedure can be restarted at the cost of losing information in the
orthogonalization process, creating the potentitial to generate new
search directions that are not orthogonal to all previous search
directions (and therefore less than optimal). From an implementation
perspective, because the operator is not required to be formed, GMRES
is significantly more flexible in its usage in that there are many
instances where various processes serve to provide the action of that
operator (e.g. radiation transport sweeps \citep{evans_2010}) that
normally may not be amenable to its full construction.

@@

We will look at the three primary parallel matrix/vector operations as
preconditioning is not an immediate requirement for implementing the
algorithms. Furthermore, variants are available that reduce the number
of global communications required (consider \cite{sosonkina_1998} as
an example of reducing global operation counts using a different
orthogonalization procedure than Arnoldi), however, we will only
consider the basic algorithms here as this handful operations can be
generalized to fit more complicated algorithms. In all of these cases,
we assume a general matrix/vector formulation that is distributed in
parallel such that both local and global knowledge of their
decomposition is available on request. Furthermore, it is assumed that
these objects are partitioned in such a way that the parallel
formulation of the operator and vectors in
Eq~(\ref{eq:linear_problem}) will be such that each parallel process
contains only a subset of the global problem and that subset forms a
local set of complete equations.

@@

\subsubsection{Parallel Vector Product}
\label{subsubsec:parallel_vector_product}
Vector product operations are used in several instances during a
Krylov iteration including vector norm computations and the
orthogonalization procedure. By definition, the vector product is a
global operation that effectively collapses two vectors to a single
value. Therefore, we cannot eliminate all global
communications. Instead, vector product operations are formulated as
\textit{global reduction operations} that are efficiently supported by
modern message passing libraries. For the dot product of two vectors
$\ve{y}$ and $\ve{x}$, a single reduction is required such that:
\begin{equation}
  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l \:,  
  \label{eq:parallel_dot_product}
\end{equation}
where the $l$ subscript denotes a local quantity, $d_l$ is the local
vector dot product, and $d_g$ is the global dot product generated by
summing the local dot products over all $p$ processes. Parallel norm
operations can be conducted with the same single reduction. Consider
the infinity norm operation:
\begin{subequations}
  \begin{gather}
    ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]\:\\
    ||x||_{\infty,g} = \max_p ||x||_{\infty,g}\:.
  \end{gather}
  \label{eq:parallel_infinity_norm}
\end{subequations}
In this form, the local infinity norm is computed over the local piece
of the vector. The reduction operation is then formed over all $p$
processes such that the global max of the vector is computed and
distributed to all processes.

\subsubsection{Parallel Matrix-Vector Multiplications}
\label{subsubsec:parallel_mat_vec_mutliply}
We finally consider matrix-vector multiplication operations by
considering sparse matrices in a compressed storage format. For these
operations, considerably more communications will be required given
that the entire global vector is required in order to compute a single
element of the product vector.

\subsubsection{Parallel Perfomance Implications for Projection Methods}
\label{subsubsec:projection_method_performance}
For very large distributed machines, global reduction operations
required at several levels of Krylov algorithms stand to reduce
scalablitity and perfomance.

>>>>

We note here several properties of this formulation and how they may
facilitate or hinder the solution of large-scale, sparse linear
problems, also noting that these properties are common among many
Krylov methods. First, from a memory perspective GMRES is efficient in
that the operator $\ve{A}$ need not be explicitly stored. Rather, only
the ability to compute the action of that operator on a vector of
valid size is required. However, these savings in memory are balanced
by the fact that the long recurrence relations used in the Arnoldi
procedure require all vectors that span the Krylov space to be
stored. If the size of these vectors becomes prohibitive, the Arnoldi
procedure can be restarted at the cost of losing information in the
orthogonalization process, creating the potentitial to generate new
search directions that are not orthogonal to all previous search
directions (and therefore less than optimal). From an implementation
perspective, because the operator is not required to be formed, GMRES
is significantly more flexible in its usage in that there are many
instances where various processes serve to provide the action of that
operator (e.g. radiation transport sweeps \citep{evans_2010}) that
normally may not be amenable to its full construction. In addition,
the minimization problem is a straight-forward least-squares problem
where $\ve{H}$ is an upper-Hessenberg matrix.

@@

We will look at the three primary parallel matrix/vector operations as
preconditioning is not an immediate requirement for implementing the
algorithms. Furthermore, variants are available that reduce the number
of global communications required (consider \cite{sosonkina_1998} as
an example of reducing global operation counts using a different
orthogonalization procedure than Arnoldi), however, we will only
consider the basic algorithms here as this handful operations can be
generalized to fit more complicated algorithms. In all of these cases,
we assume a general matrix/vector formulation that is distributed in
parallel such that both local and global knowledge of their
decomposition is available on request. Furthermore, it is assumed that
these objects are partitioned in such a way that the parallel
formulation of the operator and vectors in
Eq~(\ref{eq:linear_problem}) will be such that each parallel process
contains only a subset of the global problem and that subset forms a
local set of complete equations. The form of this partitioning is
problem dependent and often has a geometric or graph-based aspect to
its construction in order to optimize communication
patterns. Libraries such as Zoltan \citep{zoltan_2007}, provide an
implementations of such algorithms.

@@

\subsubsection{Parallel Vector Product}
\label{subsubsec:parallel_vector_product}
Vector product operations are used in several instances during a
Krylov iteration including vector norm computations and the
orthogonalization procedure. By definition, the vector product is a
global operation that effectively collapses two vectors to a single
value. Therefore, we cannot eliminate all global
communications. Instead, vector product operations are formulated as
\textit{global reduction operations} that are efficiently supported by
modern message passing libraries. For the dot product of two vectors
$\ve{y}$ and $\ve{x}$, a single reduction is required such that:
\begin{equation}
  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l \:,  
  \label{eq:parallel_dot_product}
\end{equation}
where the $l$ subscript denotes a local quantity, $d_l$ is the local
vector dot product, and $d_g$ is the global dot product generated by
summing the local dot products over all $p$ processes. Parallel norm
operations can be conducted with the same single reduction. Consider
the infinity norm operation:
\begin{subequations}
  \begin{gather}
    ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]\:\\
    ||x||_{\infty,g} = \max_p ||x||_{\infty,l}\:.
  \end{gather}
  \label{eq:parallel_infinity_norm}
\end{subequations}
In this form, the local infinity norm is computed over the local piece
of the vector. The reduction operation is then formed over all $p$
processes such that the global max of the vector is computed and
distributed to all processes.

\subsubsection{Parallel Matrix-Vector Multiplications}
\label{subsubsec:parallel_mat_vec_mutliply}
We finally consider parallel matrix-vector multiplication operations
using sparse matrices in a compressed storage format. For these
operations, more complex communication patterns will be required given
that the entire global vector is required in order to compute a single
element of the local product vector. Fortunately, the vast majority of
the global vector components will be multiplied by zero due to the
sparse matrix and therefore much of the vector can be
neglected. Instead we only require data from a handful of other
processes.

\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/linear_problem/partitioned_matrix.pdftex_t} }
  \end{center}
  \caption{Partitioned linear problem.}
  \label{fig:partitioned_linear_problem}
\end{figure}

\subsubsection{Parallel Perfomance Implications for Projection Methods}
\label{subsubsec:projection_method_performance}
For very large distributed machines, global reduction operations
required at several levels of Krylov algorithms stand to reduce
scalablitity and perfomance. Furthermore, communication between
adjacent domains may also cause a bottleneck as the number of domains
used in simulation grows. The end result is that global data must be
collected and communicated. For scaling improvement, we seek a
reduction in these types of operations.
